Carryover effects in free recall reveal how past experiences influence memories of future experiences

### 2023-01-15 paper/main.tex
quick round of edits through end of methods

\title{Carryover effects in free recall reveal how prior experiences influence
memories of new experiences}

@@

\begin{abstract} We perceive, interpet, and remember ongoing experiences
through the lens of our prior experiences. Inferring that we are one type of
situation versus another can lead us to interpret the same physical experience
differently. In turn, this can affect how we focus our attention, form
expectations of what will happen next, remember what is happening now, draw on
our prior related experiences, and so on. To study these phenomena, we asked
participants to perform simple word list learning tasks. Across different
experimental conditions, we held the set of to-be-learned words constant, but
we manipulated the orders in which the words were studied. We found that these
order manipulations affected not only how the participants recalled the ordered
lists, but also how they recalled later randomly ordered lists. Our work shows
how structure in our ongoing experiences can exert influence on how we remember
unrelated subsequent experiences. \end{abstract}

@@

The contexts we encounter help us to construct \textit{situation
models}~\citep{RangRitc12, MannEtal15} or \textit{schemas}~\citep{MasiEtal22,
BaldEtal18} that describe how experiences are likely to unfold based on our
prior experiences with similar contextual cues. For example, when we enter a
sit-down restaurant, we might expect to be seated at a table, given a menu, and
served food. Priming someone to expect a particular situation or context can
also influence how they resolve potentail ambiguities in their ongoing
experiences, including ambiguous movies and narratives~\citep{YeshEtal17}.

@@

Our understanding of how we form situation models and schemas, and how they
interact with our subjective experiences and memories, is constrained in part
by substantial differences in how we study these processes. Situation models
and schemas are most often studied using ``naturalistic'' stimuli such as
narratives and movies~\citep{ZwaaEtal95,ZwaaRadv98, NastEtal20}. In contrast,
our understanding of how we organize our memories has been most widely studied
using more traditional paradigms like free recall of random word
lists~\citep{Kaha12}. In free recall, participants study lists of items and are
instructed to recall the items in any order they choose. The orders in which
words come to mind can provide insights into how participants have organized
their memories of the studied words. Because random word lists are unstructured
by design, it is not clear if or how non-trivial situation models might apply
to these stimuli. Nevertheless, there are \textit{some} commonalities between
memory for word lists and memory for real-world experiences.

Like remembering real-world experiences, remembering words on a studied list
requires distinguishing the current list from the rest of one's experience. To
model this fundamental memory capability, cognitive scientists have posited the
existence of a special representation, called \emph{context}, that is
associated with each list. According to early
theories~\citep[e.g.][]{Este55a,AndeBowe72} context representations are
composed of many features which fluctuate from moment to moment, slowly
drifting through a multidimensional feature space. During recall, this
representation forms part of the retrieval cue, enabling us to distinguish list
items from non-list items. Understanding the role of context in memory
processes is particularly important in self-cued memory tasks, such as
\textit{free recall}, where the retrieval cue is ``context'' itself.

Over the past half-century, context-based models have enjoyed impressive
success at explaining many stereotyped behaviors observed during free recall
and other list-learning tasks~\citep{Este55a, RaaiShif80, GlenEtal83,
HowaKaha02a, SiroEtal05, KimbEtal07, PolyKaha08, PolyEtalTulv, SedeEtal08,
PolyEtal09, ShanHowa12}. These phenomena include the well-known recency and
primacy effects (superior recall of items from the end and, to a lesser extent,
from the beginning of the study list), as well as semantic and temporal
clustering effects~\citep{KahaEtal08}. The contiguity effect is an example of
temporal clustering, which is perhaps the dominant form of organization in free
recall. This effect can be seen in the tendency for people to successively
recall items that occupied neighboring positions in the study list. For
example, if a list contained the sub-sequence \textsc{``absence hollow pupil''}
and the participant recalls the word \textsc{``hollow''}, it is far more likely
that the next response will be either \textsc{``pupil''} or
\textsc{``absence''} than some other list item~\citep{Kaha96}. In addition,
there is a strong forward bias in the contiguity effect: subjects make forward
transitions (i.e., \textsc{``hollow''} followed by \textsc{``pupil''}) about
twice as often as they make backward transitions, despite an overall tendency
to begin recall at the end of the list. There are also striking effects of
semantic clustering~\citep{RomnEtal93, Bous53, BousEtal54, JenkRuss52,
MannKaha12}, whereby the recall of a given item is more likely to be followed
by recall of a similar or related item than a dissimilar or unrelated one. In
general, people organize memories for words along a wide variety of stimulus
dimensions. As captured by models like the \textit{Context Maintenance and
Retrieval Model}~\citep{PolyEtal09}, the stimulus features associated with each
word (e.g.\ the word's meaning, font size, font color, location on the screen,
size of the object the word represents, etc.) are incorporated into the
participant's mental context representation~\citep{SmitVela01, MannEtal15,
Mann20, MannEtal11, MannEtal12}. During a memory test, any of these features
may serve as a memory cue, which in turn leads the participant to recall in
succession words that share stimulus features.

@@

A key mystery is whether the sorts of situation models and schemas that people
use to organize their memories of real-world experiences might map onto the
clustering effects that reflect how people organize their memories for word
lists. On one hand, situation models and clustering effects both reflect
statistical regularities in ongoing experience. Our memory systems exploit
these regularities when generating inferences about the unobserved past and
yet-to-be-experienced future~\citep{XuEtal22, SchaTurk15, RangRitc12,
BoweEtal79, MomeEtal17}. On the other hand, the rich structure of real-world
experiences and other naturalistic stimuli that enable people to form deep and meaningful
situation models and schemas have no obvious analog in simple word lists.  Often lists
in free recall studies are explicitly \textit{designed} to be devoid of exploitable
temporal structure, for example by sorting the words in a random order~\citep{Kaha12}.

% feature-rich free recall, basic manipulation conditions, preview of findings
We designed an experimental paradigm to explore how people organize their
memories for simple stimuli (word lists) whose temporal properties change
across different ``situations,'' analogous to how the content of real-world
experiences change across different real-world situations. We asked
participants to study and freely recall a series of word lists
(Fig.~\ref{fig:exp}). Across the different conditions in the experiment, we
varied the lists' presentation orders in different ways across lists. The
studied items (words) were designed to vary along three general dimensions:
semantic (word \textit{category}, and physical \textit{size} of the referent),
lexicographic (word \textit{length} and \textit{first letter}), and visual
(font \textit{color} and the onscreen \textit{location} of each word). In our
main manipulation conditions, we asked participants to study and recall eight
lists whose items were sorted by a target feature (e.g., word category). Next,
we asked them to study and recall an additional eight lists whose items had the
same features, but that were sorted in a random temporal order. We were
interested in how these order manipulations affected participants' recall
behaviors on early (sorted) lists, as well as how order manipulations on early
lists affected recall behaviors on later (unsorted) lists. We used a series of
control conditions as a baseline; in these control conditions all of the lists
were sorted randomly, but we manipulated the presence or absence of the visual
features. Finally, in an \textit{adaptive} experimental condition we used
participants' recall behaviors on early lists to manipulate, in real-time, the
presentation orders of subsequent lists. In this adaptive condition, we sought
to identify potential commonalities within and across participants in how
people organized their memories and how those organizational tendancies affect
overall performance.

@@

Participants received course credit for enrolling in our study. We asked each
participant to fill out a demographic survey that included information about
their self-reported age, gender, ethnicity, race, education, vision, reading
impairments, medications or recent injuries, coffee consumption on the day of
testing, and level of alertness at the time of testing. All components of the
demographics survey were optional. One participant elected not to fill out any
part of the demographic survey, and all other participants report some or all
of their requested demographic information.

We aimed to run (to completion) at least 60 participants in each of the two
primary control conditions and in the adaptive condition. In all other
conditions we set a target enrollment of at least 30 participants. Because our
data collection efforts were coordinated 12 researchers and multiple testing
rooms and computers, it was not feasible for individual experimenters to know
how many participants had been run in each experimental condition until the
relevant databases were synchronized at the end of each working day. We also
over-enrolled participants for each condition to help ensure that we met our
minimum enrollment targets even if some participants dropped out of the study
prematurely or did not show up for their testing session. This led us to exceed
our target enrollments for several conditions.

Participants were assigned to experimental conditions based loosely on their
date of participation. (This aspect of our procedure helped us to more easily
synchronize the experiment databases across multiple testing computers.) Of the
490 participants who opted to fill out the demographics survey, reported ages
ranged from 17 to 31 years (mean: 19.1; standard deviation: 1.356). A total of
318 participants reported their gender as female, 170 as male, and 2
participants declined to report their gender. A total of 442 participants
reported their ethnicity as ``not Hispanic or Latino,'' 39 as ``Hispanic or
Latino,'' and 9 declined to report their ethnicity. Participants reported their
races as White (345 participants), Asian (120 participants), Black or African
American (31 participants), American Indian or Alaska Native (11 particiapnts),
Native Hawaiian or Other Pacific Islander (4 participants), Mixed race (3
participants), Middle Eastern (1 participant), and Arab (1 participant). A
total of 5 participants declined to report their race. We note that several
participants reported more than one of racial category. Participants reported
their highest degrees achieved as ``Some college'' (359 participants), ``High
school graduate'' (117 participants), ``College graduate'' (7 participants),
``Some high school'' (5 participants), ``Doctorate'' (1 participant), and
``Master's degree'' (1 participant). A total of 482 participants reported no
reading impairments, and 8 reported mild reading impairments such as mild
dyslexia. A total of 489 participants reported having normal color vision and 1
participant reported that they were color blind. A total of 482 participants
reported taking no prescription medications and having no recent injuries; 4
participants reported having ADHD, 1 reported having dyslexia, 1 reported
having allergies, 1 reported a recently torn ACL/MCL, and 1 reported a
concussion from several months prior. The participants reported consuming 0 --
3 cups of coffee prior to the testing session (mean: 0.32 cups; standard
deviation: 0.58 cups). Participants reported their current level of alertness,
and we converted their responses to numerical scores as follows: ``very
sluggish'' (-2), ``a little sluggish'' (-1), ``neutral'' (0), ``a little
alert'' (1), and ``very alert'' (2). Across all participants, the full range of
alertness levels were reported (range: -2 -- 2; mean: 0.35; standard deviation:
0.89).

We dropped from our dataset the 1 participant who reported abnormal color
vision, as well as 39 participants whose data were corrupted due to technical
failures while running the experiment or during the daily database merges. In
total, this left usable data from 452 participants, broken down by experimental
condition as follows: feature rich (67 participants), reduced (61
participants), reduced (late) (41 participants), reduced (early), (42
participants), category (30 participants), size (30 participants), length (30
participants), first letter (30 participants), color (31 participants),
location (30 participants), and adaptive (60 participants). The participant who
declined to fill out their demographic survey participated in the location
condition, and we verified verbally that they had normal color vision.

@@

Our experiment is a variant of the classic free recall paradigm that we term
\textit{feature-rich free recall}. In feature-rich free recall, participants
study 16 lists, each comprised of 16 words that vary along a number of stimulus
dimensions (Fig.~\ref{fig:exp}). The stimulus dimensions include two semantic
features related to the \textit{meanings} of the words (semantic category,
referrent object size), two lexicographic features related to the
\textit{letters} that make up the words (word length in number of letters,
identity of the word's first letter), and two visual features that are
independent of the words themselves (text color, presentation location). Each
list contains four words from each of four different semantic categories and
two object sizes; all other stimulus features are randomized. After studying
each list, the participant attempts to recall as many words as they can from
that list, in any order they choose. Because each individual word is associated
with several well-defined (and quantifiable) features, and because each list
incorporates a diverse mix of feature values along each dimension, this allows
us to evaluate participants' memory fingerprints in rich detail.

@@

Stimuli in our paradigm were 256 English words selected in a previous
study~\citep{ZimaEtal18}. The words all referred to concrete nouns, and were
chosen from 15 unique semantic categories: body parts, building-related,
cities, clothing, countries, flowers, fruits, insects, instruments,
kitchen-related, mammals, (US) states, tools, trees, and vegetables. We also
tagged each word according to the approximate size of the object the word
referred to. Words were labeled as ``small'' if the corresponding object was
likely able to ``fit in a standard shoebox'' or ``large'' if the object was
larger than a shoebox. Semantic categories varied in how many object sizes they
reflected (mean number of different sizes per category: 1.33; standard
deviation: 0.49). The numbers of words in each semantic category also varied
from 12 -- 28 (mean number of words per category: 17.07; standard deviation
number of words: 4.65). We also identified lexicographic features for each
word, including the words' first letters and lengths (i.e., number of letters).
Across all categories, all possible first letters were represented except for
`Q' (average number of unique first letters per category: 11; standard
deviation: 2 letters). Word lengths ranged from 3 -- 12 letters (average: 6.17
letters; standard deviation: 2.06 letters).

@@

The above assignments of words to lists was performed once across all
participants, such that every participant studied the same set of 16 lists. In
every condition we randomized the study order of these lists across
participants. For participants in some conditions, on some lists, we also
randomly varied two additional visual features to each word: the presentation
font color, and the word's onscreen location. These attributes were assigned
independently for word (and for every participant) at the times the words were
displayed onscreen. These visual features were varied for words in all lists
and conditions except for the ``reduced'' condition (all lists), the first
eight lists of the ``reduced (early)'' condition, and the last eight lists of
the ``reduced (late)'' condition. In these latter cases, words were all
presented in black at the center of the experimental computer's display.

To assign a random font color to each word, we selected three integers
uniformly and at random between 0 and 255, corresponding to the red (r), green
(g), and blue (b) color channels for that word. To assign random presentation
locations to each word, we selected two floating point numbers uniformly at
random (one for the word's horizontal $x$ coordinate and the other for its
vertical $y$ coordinate). The bounds of these coordinates were selected to
cover the entire visible area of the display without cutting off any part of
the words. The words were shown on 27~in (diagonal) Retina 5K iMac displays
(resolution: 5120 $\times$ 2880 pixels).

Most of the experimental manipulations we carried out entailed presenting or
sorting the presented words differently on the first eight lists participants
studied (which we call \textit{early} lists) versus on the final eight lists
they studied (\textit{late} lists). Since every participant studied exactly 16
lists, using this terminology every list was either ``early'' or ``late''
depending on its order in the list study sequence.

@@

Our experimental paradigm incorporates the Google Cloud Speech API
speech-to-text engine~\citep{HalpEtal16} to automatically transcribe
participants' verbal recalls into text. This allows recalls to be transcribed
in real time-- a distinguishing feature of the experiment; in typical verbal
recall experiments the audio data must be parsed manually. In prior work, we
used a similar experimental setup (equivalent to the ``reduced'' condition in
the present study) to verify that the automatically transcribed recalls were
sufficiently close to human-transcribed recalls to yield reliable
data~\citep{ZimaEtal18}. This real-time speech processing component of the
paradigm plays an important role in the ``adaptive'' condition of the
experiment, as described below.

@@

In the \textit{reduced (early)} condition, we followed the ``reduced''
procedure (presenting each word in black at the center of the screen) for early
lists, and followed the ``feature rich'' procedure (presenting each word in a
random color and location) for late lists. Finally, in the \textit{reduced
(late)} condition, we followed the feature rich procedure for earlylists and
the reduced procedure for late lists.

@@

Each of six \textit{order manipulation} conditions used a different
feature-based sorting procedure to order words on early lists, where each
sorting procedure relied on one relevant feature dimension. All of the
irrelevant features varied freely across words on early lists, in that we did
not consider irrelevant features in ordering the early lists. However, some
features were correlated-- for example, some semantic categories of words
referred to objects that tended to be a particular size, which means that
category and size are not fully independent. On late lists, the words were
always presented in a randomized order (chosen anew for each participant). In
all of the order manipulation conditions, we varied words' font colors and
onscreen locations, as in the feature rich condition.

\paragraph{Defining feature-based distances.} Sorting words according to a
given relevant feature requires first defining a distance function for
quantifying the dissimilarity between each pair of features. This function
varied according to the type of features. Semantic features (category and size)
are \textit{categorical}. For these features, we defined a binary distance
function: two words were considered to ``match'' (i.e., have a distance of 0)
if their labels are the same (i.e., both from the same semantic category or
both of the same size). If two words' labels were different for a given
feature, we defined the words to have a distance of 1 for that feature.
Lexicographic features (length and first letter) are \textit{discrete}. For
these features we defined a discrete distance function. Specifically, we
defined the distance between two words as either the absolute difference
between their lengths, or the absolute distance between their starting letters
in the English alphabet, respectively. For example, two words that started with
the same letter would have a ``first letter'' distance of 0, and words starting
with `J' and `A' would have a first letter distance of 9. Because words'
lengths and letters' positions in the alphabet are always integers, these
discrete distances always take on integer values. Finally, the visual features
(color and location) are \textit{continuous} and \textit{multivariate}, in that
each ``feature'' takes on multiple (positive) real values. We defined the
``color'' and ``location'' distances between two words as the Euclidean
distances between their $(r, g, b)$ color or $(x, y)$ location vectors,
respectively. Therefore the color and location distance measures always take on
positive real values (upper bounded at 441.67 for color, or 27~in for location,
reflecting the distances between the corresponding maximally different
vectors).

\paragraph{Constructing feature-sorted lists.} Given a list of words, a
relevant feature, and each word's value(s) for that feature, we developed a
stochastic algorithm for (noisily) sorting the words. First, we choose a word
uniformly at random from the set of candidates. Next, we compute the distances
between the chosen word's feature(s) and the corresponding feature(s) of all
yet-to-be-presented words. Third, we convert these distances (between the
previously presented word's feature values, $a$, and the candidate word's
feature values, $b$) to similarity scores: \begin{equation}
\mathrm{similarity}(a, b) = \exp\{-\tau \cdot \mathrm{distance}(a, b)\},
\label{eqn:similarity} \end{equation} where $\tau = 1$ in our implementation.
We note that increasing the value of $\tau$ would amplify the influence of
similarity on order, and decreasing the value of $\tau$ would diminish the
influence of similarity on order. Also note that this approach requires $\tau >
0$. Finally, we computed a set of normalized similarity values by dividing the
similarities by their sum: \begin{equation}
\mathrm{similarity}_{\mathrm{normalized}}(a, b) = \frac{\mathrm{similarity}(a,
b)}{\sum_{i=1}^n \mathrm{similarity}(a, i)}, \end{equation} where in the
demoniator, $i$ takes on each of the $n$ feature values of the to-be-presented
words. The resulting set of normalized similarity scores sums to one.

>>>>

\title{Carryover effects in free recall reveal how past experiences influence
memories of future experiences}

@@

\begin{abstract} We perceive, interpet, and remember ongoing experiences
through the lens of our prior experiences. Inferring that we are in one type of
situation versus another can lead us to interpret the same physical experience
differently. In turn, this can affect how we focus our attention, form
expectations about what will happen next, remember what is happening now, draw on
our prior related experiences, and so on. To study these phenomena, we asked
participants to perform simple word list learning tasks. Across different
experimental conditions, we held the set of to-be-learned words constant, but
we manipulated the orders in which the words were studied. We found that these
order manipulations affected not only how the participants recalled the ordered
lists, but also how they recalled later randomly ordered lists. Our work shows
how structure in our ongoing experiences can exert influence on how we remember
unrelated subsequent experiences. \end{abstract}

@@

The contexts we encounter help us to construct \textit{situation
models}~\citep{RangRitc12, MannEtal15} or \textit{schemas}~\citep{MasiEtal22,
BaldEtal18} that describe how experiences are likely to unfold based on our
prior experiences with similar contextual cues. For example, when we enter a
sit-down restaurant, we might expect to be seated at a table, given a menu, and
served food. Priming someone to expect a particular situation or context can
also influence how they resolve potential ambiguities in their ongoing
experiences, including ambiguous movies and narratives~\citep{YeshEtal17}.

@@

Our understanding of how we form situation models and schemas, and how they
interact with our subjective experiences and memories, is constrained in part
by substantial differences in how we study these processes. Situation models
and schemas are most often studied using ``naturalistic'' stimuli such as
narratives and movies~\citep{ZwaaEtal95,ZwaaRadv98, NastEtal20}. In contrast,
our understanding of how we organize our memories has been most widely studied
using more traditional paradigms like free recall of random word
lists~\citep{Kaha12, Kaha20}. In free recall, participants study lists of items and are
instructed to recall the items in any order they choose. The orders in which
words come to mind can provide insights into how participants have organized
their memories of the studied words. Because random word lists are unstructured
by design, it is not clear if or how non-trivial situation models might apply
to these stimuli. Nevertheless, there are \textit{some} commonalities between
memory for word lists and memory for real-world experiences.

Like remembering real-world experiences, remembering words on a studied list
requires distinguishing the current list from the rest of one's experience. To
model this fundamental memory capability, cognitive scientists have posited a
special context representation that is associated with each list. According to
early theories~\citep[e.g.][]{Este55a,AndeBowe72} context representations are
composed of many features which fluctuate from moment to moment, slowly
drifting through a multidimensional feature space. During recall, this
representation forms part of the retrieval cue, enabling us to distinguish list
items from non-list items. Understanding the role of context in memory
processes is particularly important in self-cued memory tasks, such as
\textit{free recall}, where the retrieval cue is ``context'' itself.
Conceptually, the same general processes might be said to describe how
real-world contexts evolve during natural experiences. However, this is still
an open area of study~\citep{Mann20, Mann21a}.

Over the past half-century, context-based models have enjoyed impressive
success at explaining many stereotyped behaviors observed during free recall
and other list-learning tasks~\citep{Este55a, RaaiShif80, GlenEtal83,
HowaKaha02a, SiroEtal05, KimbEtal07, PolyKaha08, SedeEtal08, PolyEtal09,
ShanHowa12}. These phenomena include the well-known recency and primacy effects
(superior recall of items from the end and, to a lesser extent, from the
beginning of the study list), as well as semantic and temporal clustering
effects~\citep{KahaEtal08b}. The contiguity effect is an example of temporal
clustering, which is perhaps the dominant form of organization in free recall.
This effect can be seen in the tendency for people to successively recall items
that occupied neighboring positions in the study list~\citep{Kaha96}. There are
also striking effects of semantic clustering~\citep{RomnEtal93, Bous53,
BousEtal54, JenkRuss52, MannKaha12}, whereby the recall of a given item is more
likely to be followed by recall of a similar or related item than a dissimilar
or unrelated one. In general, people organize memories for words along a wide
variety of stimulus dimensions. As formalized by models like the \textit{Context
Maintenance and Retrieval Model}~\citep{PolyEtal09}, the stimulus features
associated with each word (e.g.\ the word's meaning, font size, font color,
location on the screen, size of the object the word represents, etc.) are
incorporated into the participant's mental context
representation~\citep{SmitVela01, MannEtal15, Mann20, MannEtal11, MannEtal12}.
During a memory test, any of these features may serve as a memory cue, which in
turn leads the participant to recall in succession words that share stimulus
features.

@@

A key mystery is whether (and how) the sorts of situation models and schemas that people
use to organize their memories of real-world experiences might map onto the
clustering effects that reflect how people organize their memories for word
lists. On one hand, situation models and clustering effects both reflect
statistical regularities in ongoing experiences. Our memory systems exploit
these regularities when generating inferences about the unobserved past and
yet-to-be-experienced future~\citep{XuEtal22, SchaTurk15, RangRitc12,
BoweEtal79, MomeEtal17}. On the other hand, the rich structure of real-world
experiences and other naturalistic stimuli that enable people to form deep and meaningful
situation models and schemas have no obvious analog in simple word lists.  Often lists
in free recall studies are explicitly \textit{designed} to be devoid of exploitable
temporal structure, for example by sorting the words in a random order~\citep{Kaha12}.

% feature-rich free recall, basic manipulation conditions, preview of findings
We designed an experimental paradigm to explore how people organize their
memories for simple stimuli (word lists) whose temporal properties change
across different ``situations,'' analogous to how the content of real-world
experiences change across different real-world situations. We asked
participants to study and freely recall a series of word lists
(Fig.~\ref{fig:exp}). Across the different conditions in the experiment, we
varied the lists' presentation orders in different ways across lists. The
studied items (words) were designed to vary along three general dimensions:
semantic (word \textit{category}, and physical \textit{size} of the referent),
lexicographic (word \textit{length} and \textit{first letter}), and visual
(font \textit{color} and the onscreen \textit{location} of each word). In our
main manipulation conditions, we asked participants to study and recall eight
lists whose items were sorted by a target feature (e.g., word category). Next,
we asked them to study and recall an additional eight lists whose items had the
same features, but that were sorted in a random temporal order. We were
interested in how these order manipulations affected participants' recall
behaviors on early (sorted) lists, as well as how order manipulations on early
lists affected recall behaviors on later (unsorted) lists. We used a series of
control conditions as a baseline; in these control conditions all of the lists
were sorted randomly, but we manipulated the presence or absence of the visual
features. Finally, in an \textit{adaptive} experimental condition we used
participants' recall behaviors on early lists to manipulate, in real-time, the
presentation orders of subsequent lists. In this adaptive condition we varied
the agreement between how participants preferred to organize their memories of
the studied items versus the orders in which the items were presented.

@@

Participants received course credit for enrolling in our study. We asked each
participant to fill out a demographic survey that included questions about
their age, gender, ethnicity, race, education, vision, reading
impairments, medications or recent injuries, coffee consumption on the day of
testing, and level of alertness at the time of testing. All components of the
demographics survey were optional. One participant elected not to fill out any
part of the demographic survey, and all other participants answered some or all
of the survey questions.

We aimed to run (to completion) at least 60 participants in each of the two
primary control conditions and in the adaptive condition. In all of the other
conditions we set a target enrollment of at least 30 participants. Because our
data collection procedures entailed the coordinated efforts of 12 researchers
and multiple testing rooms and computers, it was not feasible for individual
experimenters to know how many participants had been run in each experimental
condition until the relevant databases were synchronized at the end of each
working day. We also over-enrolled participants for each condition to help
ensure that we met our minimum enrollment targets even if some participants
dropped out of the study prematurely or did not show up for their testing
session. This led us to exceed our target enrollments for several conditions.
Nevertheless, we analyze all viable data in the present paper.

Participants were assigned to experimental conditions based loosely on their
date of participation. (This aspect of our procedure helped us to more easily
synchronize the experiment databases across multiple testing computers.) Of the
490 participants who opted to fill out the demographics survey, reported ages
ranged from 17 to 31 years (mean: 19.1 years; standard deviation: 1.356 years).
A total of 318 participants reported their gender as female, 170 as male, and
two participants declined to report their gender. A total of 442 participants
reported their ethnicity as ``not Hispanic or Latino,'' 39 as ``Hispanic or
Latino,'' and nine declined to report their ethnicity. Participants reported
their races as White (345 participants), Asian (120 participants), Black or
African American (31 participants), American Indian or Alaska Native (11
particiapnts), Native Hawaiian or Other Pacific Islander (four participants),
Mixed race (three participants), Middle Eastern (one participant), and Arab
(one participant). A total of five participants declined to report their race.
We note that several participants reported more than one of racial category.
Participants reported their highest degrees achieved as ``Some college'' (359
participants), ``High school graduate'' (117 participants), ``College
graduate'' (seven participants), ``Some high school'' (five participants),
``Doctorate'' (one participant), and ``Master's degree'' (one participant). A
total of 482 participants reported no reading impairments, and eight reported
having mild reading impairments. A total of 489 participants reported having
normal color vision and one participant reported that they were red-green color
blind. A total of 482 participants reported taking no prescription medications
and having no recent injuries; four participants reported having ADHD, one reported
having dyslexia, one reported having allergies, one reported a recently torn
ACL/MCL, and one reported a concussion from several months prior. The
participants reported consuming 0 -- 3 cups of coffee prior to the testing
session (mean: 0.32 cups; standard deviation: 0.58 cups). Participants reported
their current level of alertness, and we converted their responses to numerical
scores as follows: ``very sluggish'' (-2), ``a little sluggish'' (-1),
``neutral'' (0), ``a little alert'' (1), and ``very alert'' (2). Across all
participants, the full range of alertness levels were reported (range: -2 -- 2;
mean: 0.35; standard deviation: 0.89).

We dropped from our dataset the one participant who reported having abnormal
color vision, as well as 39 participants whose data were corrupted due to
technical failures while running the experiment or during the daily database
merges. In total, this left usable data from 452 participants, broken down by
experimental condition as follows: feature rich (67 participants), reduced (61
participants), reduced (late) (41 participants), reduced (early), (42
participants), category (30 participants), size (30 participants), length (30
participants), first letter (30 participants), color (31 participants),
location (30 participants), and adaptive (60 participants). The participant who
declined to fill out their demographic survey participated in the location
condition, and we verified verbally that they had normal color vision and no
significant reading impairments.

@@

Our experiment is a variant of the classic free recall paradigm that we term
\textit{feature-rich free recall}. In feature-rich free recall, participants
study 16 lists, each comprised of 16 words that vary along a number of stimulus
dimensions (Fig.~\ref{fig:exp}). The stimulus dimensions include two semantic
features related to the \textit{meanings} of the words (semantic category,
referrent object size), two lexicographic features related to the
\textit{letters} that make up the words (word length in number of letters,
identity of the word's first letter), and two visual features that are
independent of the words themselves (text color, presentation location). Each
list contains four words from each of four different semantic categories and
two object sizes; all other stimulus features are randomized. After studying
each list, the participant attempts to recall as many words as they can from
that list, in any order they choose. Because each individual word is associated
with several well-defined (and quantifiable) features, and because each list
incorporates a diverse mix of feature values along each dimension, this allows
us to estimate which features participants are considering or leveraging in
organizing their memories.

@@

The stimuli in our paradigm were 256 English words selected in a previous
study~\citep{ZimaEtal18}. The words all referred to concrete nouns, and were
chosen from 15 unique semantic categories: body parts, building-related,
cities, clothing, countries, flowers, fruits, insects, instruments,
kitchen-related, mammals, (US) states, tools, trees, and vegetables. We also
tagged each word according to the approximate size of the object the word
referred to. Words were labeled as ``small'' if the corresponding object was
likely able to ``fit in a standard shoebox'' or ``large'' if the object was
larger than a shoebox. Semantic categories varied in how many object sizes they
reflected (mean number of different sizes per category: 1.33; standard
deviation: 0.49). The numbers of words in each semantic category also varied
from 12 -- 28 (mean number of words per category: 17.07; standard deviation
number of words: 4.65). We also identified lexicographic features for each
word, including the words' first letters and lengths (i.e., number of letters).
Across all categories, all possible first letters were represented except for
`Q' (average number of unique first letters per category: 11; standard
deviation: 2 letters). Word lengths ranged from 3 -- 12 letters (average: 6.17
letters; standard deviation: 2.06 letters).

@@

The above assignments of words to lists was performed once across all
participants, such that every participant studied the same set of 16 lists. In
every condition we randomized the study order of these lists across
participants. For participants in some conditions, on some lists, we also
randomly varied two additional visual features associated with each word: the
presentation font color, and the word's onscreen location. These attributes
were assigned independently for each word (and for every participant). These
visual features were varied for words in all lists and conditions except for
the ``reduced'' condition (all lists), the first eight lists of the ``reduced
(early)'' condition, and the last eight lists of the ``reduced (late)''
condition. In these latter cases, words were all presented in black at the
center of the experimental computer's display.

To select a random font color for each word, we drew three integers uniformly
and at random from the interval $\[0, 255\]$, corresponding to the red (r),
green (g), and blue (b) color channels for that word. To assign random
presentation locations to each word, we selected two floating point numbers
uniformly at random (one for the word's horizontal $x$ coordinate and the other
for its vertical $y$ coordinate). The bounds of these coordinates were selected
to cover the entire visible area of the display without cutting off any part of
the words. The words were shown on 27~in (diagonal) Retina 5K iMac displays
(resolution: 5120 $\times$ 2880 pixels).

Most of the experimental manipulations we carried out entailed presenting or
sorting the presented words differently on the first eight lists participants
studied (which we call \textit{early} lists) versus on the final eight lists
they studied (\textit{late} lists). Since every participant studied exactly 16
lists, every list was either ``early'' or ``late'' depending on its order in
the list study sequence.

@@

Our experimental paradigm incorporates the Google Cloud Speech API
speech-to-text engine~\citep{HalpEtal16} to automatically transcribe
participants' verbal recalls into text. This allows recalls to be transcribed
in real time-- a distinguishing feature of the experiment; in typical verbal
recall experiments the audio data must be parsed and transcribed manually. In
prior work, we used a similar experimental setup (equivalent to the ``reduced''
condition in the present study) to verify that the automatically transcribed
recalls were sufficiently close to human-transcribed recalls to yield reliable
data~\citep{ZimaEtal18}. This real-time speech processing component of the
paradigm plays an important role in the ``adaptive'' condition of the
experiment, as described below.

@@

In the \textit{reduced (early)} condition, we followed the ``reduced''
procedure (presenting each word in black at the center of the screen) for early
lists, and followed the ``feature rich'' procedure (presenting each word in a
random color and location) for late lists. Finally, in the \textit{reduced
(late)} condition, we followed the feature rich procedure for early lists and
the reduced procedure for late lists.

@@

Each of six \textit{order manipulation} conditions used a different
feature-based sorting procedure to order words on early lists, where each
sorting procedure relied on one relevant feature dimension. All of the
irrelevant features varied freely across words on early lists, in that we did
not consider irrelevant features in ordering the early lists. However, some
features were correlated-- for example, some semantic categories of words
referred to objects that tended to be a particular size, which meant that
category and size were not fully independent. On late lists, the words were
always presented in a randomized order (chosen anew for each participant). In
all of the order manipulation conditions, we varied words' font colors and
onscreen locations, as in the feature rich condition.

\paragraph{Defining feature-based distances.} Sorting words according to a
given relevant feature requires first defining a distance function for
quantifying the dissimilarity between each pair of features. This function
varied according to the type of features. Semantic features (category and size)
are \textit{categorical}. For these features, we defined a binary distance
function: two words were considered to ``match'' (i.e., have a distance of 0)
if their labels are the same (i.e., both from the same semantic category or
both of the same size). If two words' labels were different for a given
feature, we defined the words to have a distance of 1 for that feature.
Lexicographic features (length and first letter) are \textit{discrete}. For
these features we defined a discrete distance function. Specifically, we
defined the distance between two words as either the absolute difference
between their lengths, or the absolute distance between their starting letters
in the English alphabet, respectively. For example, two words that started with
the same letter would have a ``first letter'' distance of 0, and words starting
with `J' and `A' respectively would have a first letter distance of 9. Because
words' lengths and letters' positions in the alphabet are always integers,
these discrete distances always take on integer values. Finally, the visual
features (color and location) are \textit{continuous} and
\textit{multivariate}, in that each ``feature'' takes on multiple (positive)
real values. We defined the ``color'' and ``location'' distances between two
words as the Euclidean distances between their $(r, g, b)$ color or $(x, y)$
location vectors, respectively. Therefore the color and location distance
measures always take on positive real values (upper-bounded at 441.67 for
color, or 27~in for location, reflecting the distances between the
corresponding maximally different vectors).

\paragraph{Constructing feature-sorted lists.} Given a list of words, a
relevant feature, and each word's value(s) for that feature, we developed a
stochastic algorithm for (noisily) sorting the words. The stochastic aspect of
our sorting procedure enabled us to obtain unique lists for each participant.
First, we choose a word uniformly at random from the set of candidates. Next,
we compute the distances between the chosen word's feature(s) and the
corresponding feature(s) of all yet-to-be-presented words. Third, we convert
these distances (between the previously presented word's feature values, $a$,
and the candidate word's feature values, $b$) to similarity scores:
\begin{equation} \mathrm{similarity}(a, b) = \exp\{-\tau \cdot
\mathrm{distance}(a, b)\}, \label{eqn:similarity} \end{equation} where $\tau =
1$ in our implementation. We note that increasing the value of $\tau$ would
amplify the influence of similarity on order, and decreasing the value of
$\tau$ would diminish the influence of similarity on order. Also note that this
approach requires $\tau > 0$. Finally, we computed a set of normalized
similarity values by dividing the similarities by their sum: \begin{equation}
\mathrm{similarity}_{\mathrm{normalized}}(a, b) = \frac{\mathrm{similarity}(a,
b)}{\sum_{i=1}^n \mathrm{similarity}(a, i)}, \end{equation} where in the
demoniator, $i$ takes on each of the $n$ feature values of the to-be-presented
words. The resulting set of normalized similarity scores sums to one.
